Factors affecting inferences on natural mortality and associated environmental effects in state-space age-structured assessment models

### 2025-06-09 Ecov_study/mortality/manuscript/manuscript.Rmd
saving work on Ecov-M manuscript

We completed a large scale simulation study with 288 operating models, each with 100 simulated data sets, and 12 estimated models fit to each simulated data set. The factors defining operating model configuration included the source of process error on the population (recruitment, recruitment and survival, recruitment and natural mortality), the degree of temporal variation and autocorrelation of the environmental covariate, the uncertainty in the observation of the covariate, the uncertainty in indices and age composition observations, fishing history, and the magnitude of the effect of the covariate on natural mortality. The estimating models make alternative assumptions on whether to include the environmental effect, whether the mean/intercept log natural mortality (log(0.2)) is estimated or known, and whether process errors are on just recruitment, recruitment and survival, or recruitment and natural mortality.

We found convergence of all estimation models was generally best when operating models assumed process errors in recruitment and survival, constant fishing rate, greater contrast in the true environmental covariate, and lower uncertainty in corresponding observations. Reliable convergence of all estimating models also occurred with the same process errors in the operating model and a step-change in fishing, but this also required lower uncertainty in index and age composition observations. Estimating models with process errors on recruitment and survival were unlikely to converge when the process errors in the operating model did not match whereas estimating models with process errors in recruitment and natural mortality converged for operating models without this match in certain cases. Probability of convergence generally decreased when the mean/intercept log-natural mortality rate parameter was estimated.

Whether the mean log-natural mortality was estimated or not, the best accuracy of AIC for model selection occurred for models with process errors on recruitment and survival. AIC accuracy was poor for models with process errors on recruitment and natural mortality. estimating the mean natural mortality rate had small effects on the accuracy of AIC in selecting the appropriate process error. Estimating the mean log-natural mortality resulted in a small decrease in AIC accuracy. AIC was conservative for determining whether the environmental covariate affected natural mortality. AIC was very accurate in determining no effect when there was no effect in the operating model, but AIC often ranked the null model best when there was an effect. Accuracy of AIC for covariate effects improved with increased effect size, increased temporal contrast in the covariate, and lower uncertainty in observations.

We found no evidence of bias in estimation of environmental effects regardless of process error assumptions when there was low uncertainty in the environmental observations and large constrast in the environmental covariate. In most cases the relative error of the estimated environmental effect did not depend on the source of process error assumed in the estimating model. The worst bias was observed when OMs assumed R+S process errors, high uncertainty in covariate observations, low variability in the covariate, and low uncertainty in index and age composition observations. Simultaneously estimating the mean/intercept log natural mortality resulted in larger variation in the relative errors of the estimated environmental effect. Estimation of the intercept was reliable for all EM process error assumptions when the operating models assumed process errors on recruitment and natural mortality, contrast in fishing pressure over time, and lower observation error. Estimating the mean/intercept for log natural mortality generally resulted in highly variable estimates of annual natural mortality and spawning biomass and evidence of bias for some operating and estimation model assumptions about process error source. Again reliability of annual natural mortality estimates was generally improved with lower observation error uncertainty and contrast in fishing pressure.

Reliable detection of covariate effects requires informative data. AIC preferred simpler models than the true model when information content in data and contrast in covariates and abundance were low. The null model for environmental covariate effects (no covariate effect) was selected when contrast in the time series was low and/or uncertainty in observations was high. The selection of the null model by AIC also likely decreases with strength of the effect of the covariate on M. Similarly, when there was process error in recruitment and natural mortality, estimation models with process error only in recruitment were preferred presumably due to low variation in simulated natural mortality process errors.  Covariate effect estimation can be robust to process error assumptions with high contrast in covariate and low observation error.

@@

State-space stock assessment models, with non-linear functions of latent processes and numerous observation types with different probability distribution assumptions represent one of most complex classes of state-space models. The literature on the ways we make inferences and the effects of various factors on reliabilty of inferences from state-space assessment models is growing [@lietal24;@milleretal_inreview1;cadiganetal]. The importance of contrast in population size and fishing mortality and quality of data used to fit assessment models including the state-space variety is known [@magnussonhilborn07;@milleretal_inreview1]. Furthermore, estimation of natural mortality, and even temporal variablity is possible in many scenarios [@leeetal11;@johnsonetal16;@cadigan16;@millerhyun18;@milleretal_inreview1].

The effects of temporal variation in recruitment via unspecified or specified environmental factors have been extensively investigated in both traditional assessment models and state space models [@myers98;@milleretal16, haltuchpapers]. Reliability of estimating environmental and spawning biomass effects on recruitment requires a combination of strong effects, good age composition data quality, contrast in the environmental covariate and lower recruitment variability [@brittenetal_inreview;@milleretal_inreview1].

Temporal and environmental effects on growth and weight at age are also more important for short term projections and the reliability of estimation of those effects [@correaetal23; @correaetal_inreview].

However, temporal variation in natural mortality and covariate effects on natural mortality are less studied @cadigan16. State-space assessment models currently used can treat the changes in cohort abundance over time as random effects and or  despite the importance of natural mortality in inferences for the size of fish populations, their productivity and projections necessary for making catch advice. In fact, natural mortality plays a more significant role in short term projections than recruitment due to immediate effects on older age classes that constitute spawning biomass and catch.

See @milleretal_inreview1 for relevance of project 0 results (estimability of natural mortality). Some difficulty in distinguishing variation in natural mortality or effects of explicit covariates when variation in M random effects or Ecov is low relative to observation error.

@derisoetal08 formulated the same natural mortality model as a function of covariates and random effects as we use in WHAM.

 Mis-specified temporal population process errors  could lead to misidentification of stock status and biased population estimates and poor fisheries management decisions [@trijouletetal20; @legaultpalmer16; @szuwalskietal18; @croninpunt21;@liljestrandetal24].

In the present study, we conduct a simulation study with operating models (OMs) varying by degree of observation error, source and variability of process error, and fishing history. The simulations from these OMs are fitted with estimation models (EMs) that make alternative assumptions for sources of process error, whether a SRR was estimated, and whether natural mortality is estimated. Given the confounding nature of process errors, developing diagnostic tools to detect model misspecification is of great scientific interest and could aid the next generation of stock assessments [@augeretal21]. We evaluate whether convergence and Akaike Information Criterion (AIC) can correctly determine the source of process error and the existence of a SRR. We also evaluate when retrospective patterns occur and the degree of bias in the outputs of the assessment model that are important for management.

<!--
Application of state-space models in fisheries stock assessment and management has expanded dramatically within ICES, Canada and the Northeast US [@nielsenberg14;@cadigan16;@stockmiller21]. State-space approaches that use random effects to parameterize process errors is considered best practice and a requirement for the next generation of stock assessment models.

Much is known about the reliability of state-space models that are linear or Gaussian [@aeberhardetal18], but applications in fisheries management are nonlinear and typically include multiple types of observations with varying distributional assumptions. Furthermore, there is a wide range of potential random effects structures and model parameters that can be treated as random effects in assessment models. We know relatively little about the statistical reliability of such models or the ability of information criteria to distinguish among such alternative structures.

 But those studies focus primarily on Gaussian linear models.
review literature on reliability of hidden/latent process models. Primarily in other fields.

Convergence results can be useful for understanding how bad convergence in applications to real data might direct the practitioner to which alternative random effects configurations is be more appropriate. The type of convergence that we might use as a diagnostic is important. And what rule of thumb we use for certain criteria e.g. 1e-6? large SEs/CVs?
-->

Estimation of natural mortality is known to be challenging in stock assessment models (@leeetal11, others). However, estimation has been shown to be reliable in some situations [@millerhyun18]

Miller et al in review (project 0) found estimation of natural mortality was feasible when data are good and contrast in fishing pressure etc.

Let's focus results here on those situations so that we can reduce the plots in figs.

Variation in natural mortality has an immediate impact on projections unlike recruitment. Understanding this varation in M leads to better understanding of post-recruit productivity and therefore management.

Here we conduct a simulation study with OMs varying by degree of observation error uncertainty, sources of process error, fishing history, temporal variation in environmental covariates, and magnitude of the effect of the covariate on natural mortality. The simulations from these operating models are fitted with estimating models that make alternative assumptions for sources of process error, and whether (mean) M is estimated. We evaluate effects of these factors on convergence of fitted models, whether AIC can correctly determine the correct source of process error and correct assumption about covariate effects on natural mortality, and the degree of bias in relevant parameters and outputs of the assessment model.

@@

For each simulated data set we fit a set of 12 estimating models (EMs). The factors that distinguish the estimating models which are also described in detail below include source of population process error type (3 levels) whether (median) natural mortality rate was estimated or assumed known (2 levels), and whether the effect of the environmental covariate on natural mortality was estimated or not (2 levels).

The sources of population process error that were used in the OMs or assumed in the EMs were on recruitment only (R), recruitment and changes in cohort abundance over time (R+S), or recruitment and natural mortality (R+M). We did not use the log-normal bias-correction feature for process errors or observations described by @stockmiller21 for operating and EMs [@lietal_inreview]. Simulations were all carried out on the University of Massachusetts Green High-Performance Computing Cluster. Code for completing the simulations and summarizing results can be found at https://github.com/timjmiller/SSRTWG/ecov_study/mortality.

@@

Here we consider both covariate effects and process errors for natural mortality and the general model for natural mortality in year $y$ is a log-linear function of a  parameter $\beta_{M}$ that defines median natural mortality and possibly effects of the environmental covariate $X_y$ described above and autocorrelated normally distributed annual process error $\varepsilon_{M,y}$
$$
\log M_y = \beta_M + \beta_{E} X_y + \varepsilon_{M,y}
$$
$\epsilon_{M,y}|\epsilon_{M,y-1} \sim \text{N}\left(\epsilon_{M,y-1},(1-\rho_M^2)\sigma_M^2\right)$ [@stockmiller21]. We assume the median natural mortality rate $e^{\beta_M} = 0.2$ is constant across ages. For R and R+S OMs and EMs, $\epsilon_{M,y} = 0$. For all R+M OMs, we assume the same standard deviation $\sigma_M = 0.3$ and is estimated in the R+M EMs. The covariate effect is one of 3 alternative values in the operating models ($\beta_\text{E} \in \{0,0.25,0.5\}$. The parameters defining the simulated covariate time series, size of the covariate effect, and any natural mortality random effects result in a range of different levels of variation in annual natural mortality rates (Figure \ref{M_example}).

We assumed expected recruitment each year from a Beverton-Holt stock-recruit relationship (SRR)
\[
R_{y} = \frac{a \text{SSB}_{y-1}}{1 + b \text{SSB}_{y-1}}.
\]
All biological inputs to calculations of spawning biomass per recruit (i.e., weight, maturity, and natural mortality at age) are constant in the R and R+S OMs without covariate effects on natural mortality. Therefore, steepness and equilibrium unfished recruitment are also constant over the time period for those OMs [@millerbrooks21]. As in @milleretal_inreview1, our assumed biological inputs and selectivity (defined below) with constant natural mortality result in equilibrium fishing mortality that reduces spawning biomass per recruit to 40\% of the unfished level is $\Fspr[40] = 0.348$. With an assumed unfished recruitment of $R_0 = e^{10}$, setting $\Fmsy = \Fspr[40]$ results in a steepness of 0.69 and $a=0.60$ and $b = 2.4 \times 10^{-5}$.

For R+M OMs and all OMs with covariate effects on natural mortality, steepness is not constant, but we used the same $a$ and $b$ parameters as other operating models which equates to a steepness and $R_0$ at the median of the time series process for M. For operating models with time-varying random effects for fishery selectivity, \Fmsy is also not constant however we use the same fishing history as other operating models which corresponds to \Fmsy at the mean selectivity parameters.

@@

For each data set simulated from an operating model 12 estimating models were fit. There were three factors defining the configuration of each estimating model: 1) whether the mean natural mortality $\beta_M$ was estimated or assumed known, 2) whether an environmental effect $\beta_E$ was estimated or not, and 3) whether the process errors were assumed on recruitment only (R), recruitment and survival (R+S), or recruitment and natural mortality (R+M).

The configuration of the process errors in the estimating models generally matched the corresponding options in the operating models. For example, uncorrelated R+S was assumed for both the estimating and operating model. However, R+M EMs did not assume M random effects were uncorrelated ($\rho_M$ was estimated). The environmental covariate observations were included in all estimation models to ensure comparability of AIC. All fixed effects parameters for selectivity, catchability, fully-selected fishing mortality, mean recruitment, initial abundance at age, and variances for logistic-normal age composition distributions were estimated. Any process error variance parameters for recruitment, survival, and natural mortality were also estimated. The observation error variance of the environmental observations and aggregate catch and indices were all assumed known at the true values.

@@

The first measure of reliability we investigated was frequency of convergence when fitting each estimating model to the simulated data sets. There are various ways to assess convergence of the fit, but we defined successful convergence as the hessian of the marginal log-likelihood being invertible and providing variance estimates for the fixed effects parameters.

@@

We measured the frequency of correct model selection using marginal AIC. For a given operating model the set of models that were considered all made the same assumptions on whether or not to estimate (mean) natural mortality rate or it is assumed at the true value $\beta_M = log(0.2)$. For model $m$, the marginal AIC is a function of the marginal log-likelihood maximized with respect to the fixed effects in the model $\boldsymbol{\theta}$ and the number of fixed effects $n\left(\boldsymbol{\theta}\right)$ estimated,
$$
\text{AIC}_m = -2\left[{\text{argmax}}_{\boldsymbol{\theta}} \log L_m\left({\boldsymbol{\theta}}\right) - n\left({\boldsymbol{\theta}}\right)\right].
$$
All model fits that successfully completed the optimization were used in these results. We used all of these fits because some lack of convergence would be expected for the correct behavior of more complicated models that include process errors that did not exist in the operating model. For example R+M EMs fit to R OMs would be expected to estimate no variance in the natural mortality  random effects and the estimated variance parameter going to zero would cause poor convergence.

### Parameter inference reliability {-}

All results here use OM simulations with fits that satisfied the convergence criterion described above. We used this conditioning, to reflect how practitioners would proceed in analyses of model fits with real assessment data. That is, practitioners would ensure models converged such that hessian-based standard errors were available for all model parameter estimates.

We calculated median errors (ME) of $\beta_E$, the effect of environmental covariates on natural mortality, and the median natural mortality rate parameter $\beta_M$ (the mean log-natural mortality rate). We calculated the median relative errors (MRE) of terminal year estimates of natural mortality rate, spawning stock biomass, and fully-selected fishing mortality rate.
In preliminary analyses we inspected MRE of these parameters over the whole time series, but we found no appreciable differences in patterns across the various factors defining the OMs and EMs. Furthermore, results for terminal year fishing mortality were generally inversely related to those for spawning stock biomass, and are provided in the Supplementary Materials and not discussed further.

We also calculated the root mean square error (RMSE) and estimated probability of coverage of constructed 95\% confidence intervals for $\beta_E$ and $\beta_M$ for EMs that estimated these parameters. Confidence intervals for probabilities of coverage are constructed using the same methods as those for probabilities of convergence.

Results for fishing mortality rate For the $i$th simulated data the relative error for a parameter $\theta$ provided from the fitted estimation model is
$$
{\text{RE}_i}\left(\theta\right) = \frac{\widehat \theta_i - \theta_i}{\theta_i}
$$
and  constructed 95\% confidence intervals using the binomial distribution approach as in @stockmiller21 and @milleretal_inreview1. Because of the inverse relationship of the the confidence interval and the null hypothesis significance test, when the confidence interval contains zero we do not have evidence of bias from the simulation study.

For natural mortality rate, SSB, and fully-selected fishing mortality rate, we summarized results for each of the annual values, but we present results only for estimates from the terminal year, because there were no appreciable differences between the results among all years.

@@

Because @milleretal_inreview1 found inferences are most reliable for scenarios with low observation error for indices and age composition data and with temporal contrast in fishing pressure, we restrict our attention to scenarios with these characteristics in the main paper (temporal contrast in fishing and low observation error), but include corresponding results for the other scenarios in the Supplementary Materials.

@@

When the median natural mortality rate was assumed known, corresponding to usual practice in application of assessment models, good convergence for all EMs was observed for R+S operating models with \Fmsy fishing history, more variation in the latent environmental covariate ($\sigma_E = 0.5$), and lower error in the associated observations $\sigma_e = 0.1$ (Figure \ref{convergence}). Good convergence for all EMs was also observed for R+S OMs with the step change in fishing history, but with low error in indices and age comp observations.

EMs that assumed random effects just on recruitment (R), nearly always converged across all operating models that assumed more variation in the latent environmental covariate ($\sigma_E = 0.5$) and lower error in the associated observations $\sigma_E = 0.1$. EMs that assumed recruitment and survival random effects (R+S), had poor converge probability when the OMs had alternative process errors (R or R+M). The R+S EMs showed best convergence for the R+S OMs where there was lower error in the environmental observations $\sigma_E = 0.1$ or higher error in environmental observations of latent true environmental covariates with greater temporal variation. EMs that assumed recruitment and natural mortality random effects (R+M), had poor convergence probability for all OMs that had a step change in fishing history and higher uncertainty in indices and age composition.

When the OMs and EMs both assume process errors only on recruitment (R) or on recruitment and survival (R+S), convergence was worst with less variation in the true environmental covariate and larger uncertainty in associated observations. When the OMs and EMs both assume process errors on recruitment and natural mortality (R+M) convergence was problematic for all OMs with the step-change in fishing history. The best convergence was observed with this match between OMs and EMs was when the fishing history was constant and there was low uncertainty in environmental observations.

As might be expected, there was an overall drop in the probability of convergence when the mean natural mortality rate was estimated rather than assumed at the true value (Figure \ref{convergence}). Otherwise, general trends described above with mean natural mortality fixed, apply when estimated.

@@

We only present results for EMs where the median natural mortality rate parameter was estimated. Results differed little between results where this parameter was estimated or assumed known, but all of these results can be found in the Supplementary materials.

When fitting the EMs to R and R+S OMs, AIC was generally accurate in determining the correct source of process errors and, when no effect of the covariate was simulated in the OMs, AIC was also accurate for the correct treatment of the environmental covariate $\beta_E = 0$ (Figs. \ref{aic}, \ref{aic_Rom}, \ref{aic_RSom}). However, when OMs included an effect of the covariate AIC accuracy for estimating an effect in the EM was only found when the OMs also included lower covariate observation error and higher effect size and higher temporal variability in the covariate. On the other hand, for R+M Oms, AIC generally was not able to accurately determine the correct source of process errors, but was able to accurately determine the correct treatment of the environmental covariate (Figs. \ref{aic} and \ref{aic_RMom}).

When estimating models assumed the mean natural mortality rate was known, the best accuracy of AIC for model selection occurred for models with R+S process errors. R+S estimating models ranked best with very low frequency for R or R+M operating models and with very high frequency for R+S operating models (Figure \ref{aic}). R estimating models were determined best with high frequency for R operating models, but also for R+M operating models. R+M estimating models were rarely determined best for any operating models including those where the process errors matched.

AIC was conservative for determining whether the environmental covariate affected natural mortality. AIC was highly accurate in determining no effect when there was no effect in the operating model, but AIC ranked the null model best with high frequency even when there was an effect in the operating model in many cases. However the accuracy of AIC improved in certain operating models. Increased effect size, increased temporal contrast in the covariate, and lower uncertainty in all observations types lead to increased accuracy of determining covariate effects.

Relative to the assumption that the mean natural mortality rate was known, estimating the mean natural mortality rate had small effects on the accuracy of AIC in selecting the appropriate process error and whether the covariate affect natural mortality (Figure \ref{aic}). Where there were differences there were small decreases in accuracy for determining the appropriate process error and determining a covariate effect when there was one.

## Bias {-}

### Environmental effect {-}

>>>>

Treatment of natural mortality is a major consideration in assessment models and state-space approaches allow estimation of temporal variation in this mortality rate as well as effects of specific covariates. However, there has been no investigation of the reliability of inferences made regarding natural mortality, associated covariate effects and important assessment output from state-space assessment models exists. We conducted a large-scale simulation study that considers models fit to data simulated from operating models with alternative assumptions defined by several factors, but we focus on scenarios where there is temporal contrast in fishing pressure and lower uncertainty in population observations (age composition and indices of abundance). The estimating models we fit to simulated observations had alternative assumptions on whether to include the environmental effect, whether the median natural mortality rate was estimated or known, and the source of temporal variability in the population demography. Lower uncertainty in covariate observations, higher covariate temporal variability and the source of demographic temporal variability were important for better inferences regarding the covariate effect on natural mortality.

**keywords:** state-space assessment models, time-varying natural mortality, bias, AIC

@@

State-space stock assessment models, with non-linear functions of latent processes and numerous observation types with different probability distribution assumptions represent one of most complex classes of state-space models. The literature on the effects of various factors on reliabilty of inferences from state-space assessment models is growing [@lietal24;@milleretal_inreview1]. The importance of contrast in population size and fishing mortality and quality of data used to fit assessment models including the state-space variety is known [@magnussonhilborn07;@milleretal_inreview1]. Furthermore, estimation of natural mortality, and even temporal variablity is possible in many scenarios [@leeetal11;@cadigan16;@millerhyun18;@milleretal_inreview1].

The effects of temporal variation in recruitment via undefined or explicit environmental factors have been extensively investigated in both traditional assessment models and state space models [@myers98; @haltuchpunt11; @johnsonetal16; @milleretal16]. Reliability of estimating environmental and spawning biomass effects on recruitment in state-space assessment models requires a combination of strong effects, good age composition data quality, contrast in the environmental covariate and lower recruitment variability [@brittenetal_inreview;@milleretal_inreview1].

A critical aspect of fisheries assessment models and their use in management is short-term projections that are used to determine catch advice. While understanding drivers of recruitment is important particularly for subsequent effects on reference points, recruitment in short-term projections typically has little impact on the exploitable biomass in the first few projection years. However, assumptions for natural mortality ($M$) have immediate and larger effects on projected biomass because they affect the abundances at older age classes at the end of the data time series that constitute spawning biomass and catch [@brodziaketal08;@stocketal21].

Because of the effects of $M$ on both biological reference points and short term projections, better understanding sources of variation in $M$ would provide more accurate estimation of abundance and productivity and therefore improved management. Temporal variation in $M$ is less studied than recruitment, but its importance for explaining variability in observations has been demonstrated in state-space assessment models for Atlantic cod and yellowtail flounder [@cadigan16;@stocketal21]. @derisoetal08 also demonstrated the importance of several factors affecting Pacific herring natural mortality.

Assessment models could include temporal variation in many aspects of population dynamics or how observations are related to the population. For example the Woods Hole Assessment Model (WHAM) can include process errors treated as random effects for transition in cohorts over time, catchability for indices of abundance, selectivity of fishing fleets or indices, movement between regions, or in $M$ [@stockmiller21;@milleretal_inreview]. However, misspecified temporal population process errors could lead to biased population and stock status estimation, and, therefore, poor fisheries management decisions [ @legaultpalmer16; @szuwalskietal18]. Studies of the reliability of inferences we might make regarding the presence of temporal variability in $M$ are limited. @milleretal_inreview1 found AIC could accurately distinguish process errors in the changes in abundance for cohorts, but not for those specifically due to $M$ except when uncertainty in population observations (indices, catch and age composition) was low and there was greater temporal variation in natural mortality. In their simulation studies looking a models with multiple sources of process error, @lietal24 found including more sources of process error than existed in the operating model was a better model-building approach than excluding them a priori.

Here we conduct a simulation study with operating models (OMs) varying by degree of observation error uncertainty, sources of process error, fishing history, temporal variation in environmental covariates, and magnitude of the effect of the covariate on natural mortality. The simulated observations from these OMs are fitted with estimating models that make alternative assumptions for sources of process error, and whether median M and covariate effects are estimated. We evaluate the effects of these factors on convergence of fitted models, whether Akaike's information criterion (AIC) can correctly determine the correct source of process error and correct assumption about covariate effects on natural mortality, and the degree of bias in inferences for relevant parameters and outputs of the assessment model.

@@

For each simulated data set we fit a set of 12 estimating models (EMs). The factors that distinguish the estimating models which are also described in detail below include source of population process error type (3 levels) whether (median) $M$ was estimated or assumed known (2 levels), and whether the effect of the environmental covariate on $M$ was estimated or not (2 levels).

The sources of population process error that were used in the OMs or assumed in the EMs were on recruitment only (R), recruitment and changes in cohort abundance over time (R+S), or recruitment and $M$ (R+M). We did not use the log-normal bias-correction feature for process errors or observations described by @stockmiller21 for operating and EMs [@lietal_inreview]. Simulations were all carried out on the University of Massachusetts Green High-Performance Computing Cluster. Code for completing the simulations and summarizing results can be found at https://github.com/timjmiller/SSRTWG/ecov_study/mortality.

@@

The general model for $M$ in year $y$ is a log-linear function of both covariate effects $X_y$ and process errors $\varepsilon_{M,y}$ and a  parameter $\beta_{M}$ that defines median $M$ 
$$
\log M_y = \beta_M + \beta_{E} X_y + \varepsilon_{M,y}
$$
where the process errors are modeled as random effects that may, in general, be autocorrelated normal random variables 
$$
\varepsilon_{M,y}|\varepsilon_{M,y-1} \sim \text{N}\left(\varepsilon_{M,y-1},(1-\rho_M^2)\sigma_M^2\right)
$$ 
[@stockmiller21], but we assume $\rho_M = 0$ in our R+M OMs. We assume the median $M$ rate $e^{\beta_M} = 0.2$ is constant across ages. For R and R+S OMs and EMs, $\varepsilon_{M,y} = 0$. For all R+M OMs, we assume the same standard deviation $\sigma_M = 0.3$ and is estimated in the R+M EMs. The covariate effect is one of 3 alternative values in the operating models, $\beta_\text{E} \in \{0,0.25,0.5\}$. The parameters defining the simulated covariate time series, size of the covariate effect, and any $M$ random effects result in a range of different levels of variation in annual values (Figure \ref{M_example}).

We assumed expected recruitment each year is from a Beverton-Holt stock-recruit relationship (SRR)
\[
R_{y} = \frac{a \text{SSB}_{y-1}}{1 + b \text{SSB}_{y-1}}.
\]
All biological inputs to calculations of spawning biomass per recruit (i.e., weight, maturity, and $M$ at age) are constant in the R and R+S OMs without covariate effects on $M$. Therefore, steepness and equilibrium unfished recruitment are also constant over the time period for those OMs [@millerbrooks21]. As in @milleretal_inreview1, our assumed biological inputs and selectivity (defined below) with constant $M$ result in equilibrium fishing mortality that reduces spawning biomass per recruit to 40\% of the unfished level is $F_{40\%} = 0.348$. With an assumed unfished recruitment of $R_0 = e^{10}$, setting $\Fmsy = F_{40\%}$ results in a steepness of 0.69 and $a=0.60$ and $b = 2.4 \times 10^{-5}$. For R+M OMs and all OMs with covariate effects on $M$, steepness is not constant, but we used the same $a$ and $b$ parameters as other operating models which equates to a steepness and $R_0$ at the median of the time series processes for $M$ and the covariate.

@@

For each data set simulated from an operating model 12 estimating models were fit. There were three factors defining the configuration of each estimating model: 1) whether $\beta_M$ was estimated or assumed known, 2) whether an environmental effect $\beta_E$ was estimated or not, and 3) whether the process errors were assumed on recruitment only (R), recruitment and survival (R+S), or recruitment and $M$ (R+M).

The configuration of the process errors in the estimating models generally matched the corresponding options in the operating models. For example, uncorrelated R+S was assumed for both the estimating and operating model. However, R+M EMs did not assume M random effects were uncorrelated ($\rho_M$ was estimated). The environmental covariate observations were included in all estimation models to ensure comparability of AIC. All fixed effects parameters for selectivity, catchability, fully-selected fishing mortality, mean recruitment, initial abundance at age, and variances for logistic-normal age composition distributions were estimated. Any process error variance parameters for recruitment, survival, and $M$ were also estimated. The observation error variance of the environmental observations and aggregate catch and indices were all assumed known at the true values.

@@

We measured the frequency of convergence when fitting each EM to the simulated data sets for each OM. There are various ways to assess convergence of the fit [e.g., @carvalhoetal21], but we defined successful convergence as the Hessian of the marginal log-likelihood being invertible and providing variance estimates for the fixed effects parameters as recommended by @milleretal_inreview1. We calculated 95\% confidence intervals for probability of convergence using the Clopper-Pearson exact method [@clopperpearson34;@thulin14].

@@

We measured the frequency of correct model selection using marginal AIC. For a given operating model the set of models that were considered all made the same assumptions on whether or not to estimate $\beta_M$ or it is assumed at the true value. For model $m$, the marginal AIC is a function of the marginal log-likelihood maximized with respect to the fixed effects in the model $\boldsymbol{\theta}$ and the number of fixed effects $n\left(\boldsymbol{\theta}\right)$ estimated,
$$
\text{AIC}_m = -2\left[{\text{argmax}}_{\boldsymbol{\theta}} \log L_m\left({\boldsymbol{\theta}}\right) - n\left({\boldsymbol{\theta}}\right)\right].
$$
All model fits that successfully completed the optimization were used in these results. We used all of these fits because some lack of convergence would be expected for the correct behavior of more complicated models that include process errors that did not exist in the operating model. For example R+M EMs fit to R OMs would be expected to estimate no variance in the $M$  random effects and the estimated variance parameter going to zero would cause poor convergence.

### Parameter estimation {-}

All results here use OM simulations with fits that satisfied the convergence criterion described above. We used this conditioning, to reflect how practitioners would proceed in analyses of model fits with real assessment data. That is, practitioners would ensure models converged such that Hessian-based standard errors were available for all model parameter estimates.

We focused on statistical behavior of estimators of the covariate effect on $M$ ($\widehat \beta_E$), the estimator of the median $M$ parameter ($\widehat \beta_M$), and estimators of terminal year natural mortality $\widehat M$, spawning stock biomass $\widehat{\text{SSB}}$, and fully-selected fishing mortality rate $\widehat F$. In preliminary analyses we results of the estimators of all the annual values for $M$, SSB and fishing mortality over the whole time series, but we found no appreciable differences in patterns across the various factors defining the OMs and EMs. Furthermore, results for terminal year fishing mortality were generally inversely related to those for spawning stock biomass, and are provided in the Supplementary Materials and not discussed further.

We calculated median errors (ME) of $\widehat \beta_E$, $\widehat \beta_M$), and the Hessian-based standard error estimators of these parameters ($\widehat{SE}\left(\widehat \beta_E\right)$ and $\widehat{SE}\left(\widehat \beta_M\right)$). We calculated the median relative errors (MRE) of terminal year $\widehat M$, $\widehat{\text{SSB}}$, and $\widehat F$. We also calculated the root mean square error (RMSE) and estimated probability of coverage of constructed 95\% confidence intervals for $\widehat \beta_E$ and $\widehat \beta_M$ for EMs that estimated these parameters. We constructed the confidence intervals for probabilities of CI coverage using the same methods as those for probabilities of convergence.

The true values for terminal year SSB and fishing mortality rate vary among simulations.  For the $i$th simulated data the relative error for terminal year value $\theta_i$ provided from the fitted estimation model is
$$
\text{RE}_i\left(\theta\right) = \frac{\widehat \theta_i - \theta_i}{\theta_i}
$$
We calculated RMSE as
$$
\text{RMSE}\left(\widehat \theta\right) = \sqrt{\frac{1}{n}\sum^n\left(\widehat \theta_i - \theta_i\right)^2}
$$
For ME and RME of estimators, we constructed 95\% confidence intervals using the binomial distribution approach as in @stockmiller21 and @milleretal_inreview1. Because of the inverse relationship of the the confidence interval and the null hypothesis significance test, when the confidence interval contains zero we do not have evidence of bias from the simulation study.

@@

@milleretal_inreview1 found inferences are most reliable for distinguishing process error sources and stock-recruit relationships in scenarios with lower observation error for indices and age composition data and with temporal contrast in fishing pressure. Our expectation is that inferences for covariate effects on $M$ would require data to be at least that informative and inspection of the comprehensive set of results across all OMs generally confirms this (See supplemental Materials). Therefore, we restrict our attention to these OM scenarios with more informative data in the main paper (temporal contrast in fishing and lower population observation error).

@@

In OMs with contrast in fishing pressure and lower uncertainty in population observations (catch, indices and age composition), R EMs generally converged with high frequency except when EMs estimated covariate effects on $M$ and the covariate uncertainty ($\sigma_e$) was high (Figure \ref{convergence}). R+S EMs generally converged with low frequency when the OM process error did not match the EM configuration. R+M EMs generally did not converge with high frequency except for R+S OMs with low covariate uncertainty and where the EMs assumed median $M$ as known. There was no strong effect of the treatment of median M (known or estimated) on convergence for any of the other EMs. The largest effect of the size of the true covariate effect ($\beta_E$) on convergence was observed for R and R+M EMs that estimated covariate effects in R and R+M OMs with high uncertainty in covariate observations ($\sigma_e$) and high variability in covariate process errors ($\sigma_E$). There was an increase in convergence frequency with increased variability in covariate process errors for many EMs in OMs with large covariate uncertainty.

@@

We only present results for EMs where the median $M$ rate parameter was estimated because results differed little whether this parameter was estimated or assumed known (see Figures \ref{aic_Rom} to \ref{aic_RMom}). When there is temporal contrast in fishing pressure and lower uncertainty in population observations, the only OM scenarios where AIC was consistently accurate for both covariate effect and process error configurations were R OMs with high variability in the covariate and low uncertainty in observations of the covariate (Figure \ref{aic}). R+M OMs with those same covariate attributes were only accurate for the covariate effect (low Type I and II errors) and the incorrect process error almost always chosen was R rather than R+S. We also observed high accuracy for both covariate effect and process error configurations for R+S OMs where either no effect of the covariate or the strongest effect of the covariate was simulated. All R and R+S OMs were accurate for process error configuration and across all OMs, AIC accurately selects models without covariate effects when there are none (low Type I error). See Figures \ref{aic_Rom} to \ref{aic_RMom} for full results.

## Covariate effect estimation {-}

When there is temporal contrast in fishing pressure, lower uncertainty in population observations, low uncertainty in environmental observations, and larger temporal contrast in the simulated true environmental covariate, we observed generally little or no bias in estimation of covariate effects ($\widehat \beta_E$) across all EM and OM process error assumptions, both EM configurations of median $M$, and all true covariate effect sizes (Figure \ref{beta_E_bias}, rows 3 and 4). The only exception to this was when R EMs with median $M$ estimated were fit to R+S OMs with those configurations. Decreasing trends in bias of $\widehat \beta_E$ for EMs fit to many configurations of R and R+M OMs indicate that $\widehat \beta_E$ values were closer to zero on average even when the true effect increased. For R+S OMs, EMs with the matching process error configuration showed little evidence of bias in estimating $\beta_E$ across a range of covariate and covariate observation configurations.

Bias of Hessian-based standard error estimation for $\widehat \beta_E$ was also close to zero in the same OM scenarios where bias in estimation of $\beta_E$ itself was close to zero (Figure \ref{se_beta_E_bias}, rows 3 and 4). However, in these high information OMs bias of standard error estimation was observed in for several EMs that did not have the matching R+S and R+M process error configuration. Estimation of standard errors was essentially unreliable for most EMs when OMs had higher uncertainty in covariate observations (Figure \ref{se_beta_E_bias}, rows 5 to 8).

Despite the reliable estimation of covariate effects and standard errors with low uncertainty in environmental observations, and larger temporal contrast in the simulated true environmental covariate, we found poor confidence interval coverage for many EMs fitted to R+S and R+M OMs (Figure \ref{beta_E_CI_coverage}, rows 3 and 4). For example, R+M EMs fit to R+M OMs showed little bias in estimation of the covariate effect and standard errors, but confidence interval coverage was negatively biased (rows 3 and 4 in right two columns of Figures \ref{beta_E_bias}, \ref{se_beta_E_bias}, and \ref{beta_E_CI_coverage}. However, there appears to be a positive correlation of the covariate and standard error estimates such that estimates less than the true value had negatively biased standard error estimates which would make confidence intervals too small (Figure \ref{ex_lm_beta_E_SE_beta_E}). Confidence interval coverage was most reliable for R+S OMs when the EMs had the matching process error configuration (Figure \ref{beta_E_CI_coverage}, middle 2 columns).

## Median natural mortality rate estimation {-}
